A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-24 paper.tex
Add Prospective vs Retrospective section

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Inferring the detailed genetic genealogy of a set of sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as inputs have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we define a concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
to avoid confusion, we refer to as a ``genome ARG``, or gARG.
(We contrast this with the classical definition derived from the
coalescent process in later sections.)
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
% NOTE: does this example really help? Rather than spelling out what
% we can already see, perhaps we should point out the non-obvious
% thing that we have decoupled from "events"
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal  and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are independent recombinants of
the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}, and regions of genome transmitted are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding genome ARG along with annotations on all edges;
annotations showing partial
genome transmission are highlighted in bold. Recombination (e.g.,
the cause of the breakpoint at position 7)
is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

@@

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell). The edges are ``annotated''
with the genome coordinates over which inheritance occurs.
There are two different ways in which we use these inheritance
annotations, and the distinction is subtle and important. The difference
is illustrated in Fig~\ref{fig-ancestry-resolution}.

% TODO majorly condense this caption, it's much too long.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
TODO DELETE A? Or keep it? Maybe this would help some hard-core eARGers?
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
% (A) A conventional Griffiths eARG. Square nodes represent events, with
% each recombination (red) annotated with the position of a single breakpoint
% along the 7-unit-long chromosome.
(A) A gARG with edges annotated by . Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways to the samples. Nodes
\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
to events which have no impact on the genetic ancestry of the samples.
}
\end{figure}

The simplest interpretation of the inheritance intervals $I$ for
an ancestor genome $v$ and descendant $u$, is that $I$ defines the
regions of genome that $u$ inherited from $v$. For example,
node $o$ is the recombinant descendant of nodes $p$ and $q$ in
Figure~\ref{fig-ancestry-resolution}A, which shows that it in inherited
everything to the left of position $1$ from $p$ and the rest of
its genome from node $q$.
Thus, these edge annotations are concerned only with the genetic relationship
between the nodein question, and are not affected by the rest of the graph.
We will thefore refer to inheritance annotations of this type as ``local inheritance
annotations''. Graphs of this type result from forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree} or the
``Big ARG'' stochastic process (Appendix XXX).

ARGs are most often thought of as a \emph{retrospective} structure however,
defined with respect to some set of (usually, contemporary) sample genomes.
In this case we are not explicitly interested in the
entire interval of genome descendant $u$ inherited from ancestor $v$,
but rather the subset of that interval that is \emph{ancestral to our
sample}.
Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
reduce the annotation intervals to those that are ancestral to the
sample (note that node $q$ is now ommitted because the MRCA of the sample
has already been found for the full genome at node $p$).
Tracking such
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
is much more useful~\citep{mcgill2013graphml,arenas2013importance}
and, indeed, all that is possible to
have information about when inferring ARGs from contemporary samples.
We refer to a gARG with inheritance annotations of this type as
``sample resolved''.
% Keeping track of resolved ancestral material is one of the distinguishing
% features of ARGs vs phylogenetic
% networks~\citep{mcgill2013graphml,arenas2013importance}, but
% is actually incompatible with the Griffiths gARG encoding (it does this
% implicitly, I guess)

>>>>

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the detailed genetic genealogy of a set of sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as inputs have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
% TODO update when natgen paper comes out
nowbandegani2022extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we define a concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but, crucially,
where each individual is represented by two nodes (for diploids).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal  and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are independent recombinants of
the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
% Recombination (e.g.,
% the cause of the breakpoint at position 7)
% is associated with two parent genomes (\textsf{e} and
% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
% (C) The edge-annotated genome ARG in (B) fully defines the
(C) The corresponding local (``marginal'') trees.
}
\end{figure}

The edges in a gARG specify that genetic inheritance occurred between an
ancestor and a descendant, but the connectivity information alone
does not tell us which \emph{parts} of their genomes were inherited.

@@

% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
\section*{Prospective vs restrospective ARGs}
ARGs are usually thought of as a restrospective structure, in which we
reason about the genomes that are genetic ancestors of some set of
sample genomes. Figure~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic inheritance must have occured between (e.g.)
$I_8$ and her daughter $I_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
we have no information about it, given these sampled genomes.
Event ARGs, given their roots in the coalescent with recombination,
are inherently retrospective. They define the evolutionary
events that have occured in the genetic ancestry of a sample.
% This is weak, rephrase.
An event ARG must have some set of samples, or there
will be nothing to define events in terms of.

The gARG representation, on the other hand,
can be either retrospective or \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell),
and are annotated
with the genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these  inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
direct genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

Prospective ARGs occur in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'').
Describing this instead as a ``prospective ARG'', and emphasising
the use of the same data structures and direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm which converts a prospective
ARG into a retrospective one is a powerful tool,
with other applications as we discuss in the following sections.

\section*{Ancestry resolution}\label{Ancestry_resolution}

These genomic intervals, directly ancestral to the samples
and  scattered across a set of lineages at some time in the past,
are usually known as the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

% TODO majorly condense this caption, it's much too long.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
% (A) A conventional Griffiths eARG. Square nodes represent events, with
% each recombination (red) annotated with the position of a single breakpoint
% along the 7-unit-long chromosome.
(A) A gARG with edges annotated by . Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways to the samples. Nodes
\textsf{g},  \textsf{j}, \textsf{o} and \textsf{q} are redundant in that they correspond
to events which have no impact on the genetic ancestry of the samples.
}
\end{figure}

ARGs are most often thought of as a \emph{retrospective} structure however,
defined with respect to some set of (usually, contemporary) sample genomes.
In this case we are not explicitly interested in the
entire interval of genome descendant $u$ inherited from ancestor $v$,
but rather the subset of that interval that is \emph{ancestral to our
sample}.
Fig~\ref{fig-ancestry-resolution}B shows the graph that we obtain when
reduce the annotation intervals to those that are ancestral to the
sample (note that node $q$ is now ommitted because the MRCA of the sample
has already been found for the full genome at node $p$).
Tracking such
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
is much more useful~\citep{mcgill2013graphml,arenas2013importance}
and, indeed, all that is possible to
have information about when inferring ARGs from contemporary samples.
We refer to a gARG with inheritance annotations of this type as
``sample resolved''.

% Keeping track of resolved ancestral material is one of the distinguishing
% features of ARGs vs phylogenetic
% networks~\citep{mcgill2013graphml,arenas2013importance}, but
% is actually incompatible with the Griffiths gARG encoding (it does this
% implicitly, I guess)
